Evidence accumulation modelling offers new insights into the cognitive mechanisms that underlie linguistic and action-based training

### 2025-02-03 manuscript/manuscript.Rmd
updated the manuscript and supps to reflect editor feedback

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study [@simmons2012]. In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: [https://aspredicted.org/6ztj-45s5.pdf](https://aspredicted.org/6ztj-45s5.pdf)). We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

@@

We applied the LBA model to a previously published dataset from Cross and colleagues (2012). This study used fMRI to examine brain activity whilst participants completed a speeded perceptual discrimination task before and after training. The data used in the current study, therefore, were collected in or before 2012. Training consisted of participants learning to name 10 knots, tie 10 knots and name and tie a further 10 knots, while a final 10 knots remained untrained (i.e., no information about how to tie the knot or what it was called was experienced during the training period). Before and after the knot training portion of the study, participants were required to complete 80 trials of a perceptual discrimination task. In this task, participants had to decide whether two photographs were of the same or different knots. Each pair of knots was selected from the same learning class (i.e., name, tie, both name and tie or untrained knots). When photographs were of the same knot, the viewing angle of the two photographs differed, so the stimuli presented were never identical. Of the 80 trials, 50% of the knots were the same and 50% of the knots were different. Both accuracy and response times (ms) of the perceptual discrimination task were recorded. As we were interested in modelling only the behavioural results of the perceptual discrimination task, we summarise the aspects of the dataset relevant to the present study below and refer readers to the original paper for all other details (see Cross et al., 2012).

The data included in the present paper differed from that reported by Cross and colleagues (2012) in a number of key ways. First, while 30 participants took part in the behavioural portion of the original study, only data from the 20 participants who comprised the final fMRI sample were available for re-analysis. The decision to include only the final training session in the LBA model meant that a further two participants were excluded from modelling as they did not complete the final session of the perceptual discrimination task. Therefore, 18 participants were included in the present study (M = 19.4 years, age ranged from 17 – 27 years). Second, unlike Cross and colleagues who included the post-training session across all five days in their analysis of accuracy and RT, we included only the post-training session on the final day of training (Day 5 post-training). Given the aim of our study was to use evidence accumulation modelling to examine the cognitive mechanisms underlying linguistic and action-based training, it was sufficient to use the training effects evident on the last day of training for our analysis. That is, general training effects could be compared via a comparison between the untrained and trained conditions, and between each type of training category (e.g. linguistic vs. action vs. both). The benefit of including only the final post-training sessions was a computationally simpler model.

@@

>>>>

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study [@simmons2012]. In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: [https://aspredicted.org/6ztj-45s5.pdf](https://aspredicted.org/6ztj-45s5.pdf)). Given that we used secondary data in this study, we did not have a stopping rule in terms of data collection because we did not collect any data. Instead, we simply aimed to use as much of the existing data as possible basd on availability.

We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

@@

We applied the LBA model to a previously published dataset from Cross and colleagues (2012). This study used fMRI to examine brain activity whilst participants completed a speeded perceptual discrimination task before and after training. The data used in the current study, therefore, were collected in 2006. Training consisted of participants learning to name 10 knots, tie 10 knots and name and tie a further 10 knots, while a final 10 knots remained untrained (i.e., no information about how to tie the knot or what it was called was experienced during the training period). Before and after the knot training portion of the study, participants were required to complete 80 trials of a perceptual discrimination task. In this task, participants had to decide whether two photographs were of the same or different knots. Each pair of knots was selected from the same learning class (i.e., name, tie, both name and tie or untrained knots). When photographs were of the same knot, the viewing angle of the two photographs differed, so the stimuli presented were never identical. Of the 80 trials, 50% of the knots were the same and 50% of the knots were different. Both accuracy and response times (ms) of the perceptual discrimination task were recorded. As we were interested in modelling only the behavioural results of the perceptual discrimination task, we summarise the aspects of the dataset relevant to the present study below and refer readers to the original paper for all other details (see Cross et al., 2012).

The data included in the present paper differed from that reported by Cross and colleagues (2012) in a number of key ways. First, while 30 participants took part in the behavioural portion of the original study, only data from the 20 participants who comprised the final fMRI sample were available for re-analysis. The decision to include only the final training session in the LBA model meant that a further two participants were excluded from modelling as they did not complete the final session of the perceptual discrimination task. Therefore, 18 participants, all right-handed (Oldfield, 1971) and neurologically healthy undergraduate and graduate students, were included in the present study (M = 19.4 years, age ranged from 17 – 27 years). We do not have access to information on the gender of participants, but we note from the original publication that of the 20 participants that completed the fMRI component of the study, 14 were female (Cross et al., 2012). This means that our sample of 18 participants is predominantly female.

Second, unlike Cross and colleagues who included the post-training session across all five days in their analysis of accuracy and RT, we included only the post-training session on the final day of training (Day 5 post-training). Given the aim of our study was to use evidence accumulation modelling to examine the cognitive mechanisms underlying linguistic and action-based training, it was sufficient to use the training effects evident on the last day of training for our analysis. That is, general training effects could be compared via a comparison between the untrained and trained conditions, and between each type of training category (e.g. linguistic vs. action vs. both). The benefit of including only the final post-training sessions was a computationally simpler model.

@@

Similarly, while we had originally planned to model the difference between the pre- and post-training data for each participant, due to model complexity we were limited to including the post-training session only. It is possible, therefore, that improved post-training scores in the trained conditions, rather than reflecting the operation of training, may instead be due to unbalanced pre-test scores between these conditions. However, it should be noted that there were no clear and obvious differences in RT or accuracy in the day 1, session 1 data (i.e., before any training had started; see supplementary materials). Therefore, we consider it quite unlikely that random variation in pre-training differences between conditions could account for our findings.

In using secondary data analysis in the present study, we were also limited in what data was available for analysis. Indeed, we determined our sample size by aiming to use as much of the existing data as was possible. Consequently, the likely precision of our parameter estimates is unknown. Given this limitation, it’s important for future studies look to replicate the findings in the present study.

### 2025-02-03 manuscript/manuscript.tex
updated the manuscript and supps to reflect editor feedback

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study (Simmons, Nelson, \& Simonsohn, 2012). In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: \url{https://aspredicted.org/6ztj-45s5.pdf}). We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

@@

We applied the LBA model to a previously published dataset from Cross and colleagues (2012). This study used fMRI to examine brain activity whilst participants completed a speeded perceptual discrimination task before and after training. The data used in the current study, therefore, were collected in or before 2012. Training consisted of participants learning to name 10 knots, tie 10 knots and name and tie a further 10 knots, while a final 10 knots remained untrained (i.e., no information about how to tie the knot or what it was called was experienced during the training period). Before and after the knot training portion of the study, participants were required to complete 80 trials of a perceptual discrimination task. In this task, participants had to decide whether two photographs were of the same or different knots. Each pair of knots was selected from the same learning class (i.e., name, tie, both name and tie or untrained knots). When photographs were of the same knot, the viewing angle of the two photographs differed, so the stimuli presented were never identical. Of the 80 trials, 50\% of the knots were the same and 50\% of the knots were different. Both accuracy and response times (ms) of the perceptual discrimination task were recorded. As we were interested in modelling only the behavioural results of the perceptual discrimination task, we summarise the aspects of the dataset relevant to the present study below and refer readers to the original paper for all other details (see Cross et al., 2012).

The data included in the present paper differed from that reported by Cross and colleagues (2012) in a number of key ways. First, while 30 participants took part in the behavioural portion of the original study, only data from the 20 participants who comprised the final fMRI sample were available for re-analysis. The decision to include only the final training session in the LBA model meant that a further two participants were excluded from modelling as they did not complete the final session of the perceptual discrimination task. Therefore, 18 participants were included in the present study (M = 19.4 years, age ranged from 17 -- 27 years). Second, unlike Cross and colleagues who included the post-training session across all five days in their analysis of accuracy and RT, we included only the post-training session on the final day of training (Day 5 post-training). Given the aim of our study was to use evidence accumulation modelling to examine the cognitive mechanisms underlying linguistic and action-based training, it was sufficient to use the training effects evident on the last day of training for our analysis. That is, general training effects could be compared via a comparison between the untrained and trained conditions, and between each type of training category (e.g.~linguistic vs.~action vs.~both). The benefit of including only the final post-training sessions was a computationally simpler model.

@@

>>>>

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study (Simmons, Nelson, \& Simonsohn, 2012). In addition, the hypotheses and planned analysis for this paper were preregistered in advance (see our pre-registration here: \url{https://aspredicted.org/6ztj-45s5.pdf}). Given that we used secondary data in this study, we did not have a stopping rule in terms of data collection because we did not collect any data. Instead, we simply aimed to use as much of the existing data as possible basd on availability.

We deviated from our pre-registration in one way. Originally, we had intended to include in our model data from both a pre- and post-training session. That is, we intended to compute the difference in parameter estimates between the pre- and post-training variable. Instead, we only included data from the post-training session in our final analysis. This deviation from preregistration was necessary as the model building process could not be completed with both pre- and post-training session variables included, due to the large number of parameters being fit and the model failing to converge. By removing the pre-training session data, the number of parameters was reduced by half. As a result, the model could converge and we were able to successfully complete the model building process.

@@

We applied the LBA model to a previously published dataset from Cross and colleagues (2012). This study used fMRI to examine brain activity whilst participants completed a speeded perceptual discrimination task before and after training. The data used in the current study, therefore, were collected in 2006. Training consisted of participants learning to name 10 knots, tie 10 knots and name and tie a further 10 knots, while a final 10 knots remained untrained (i.e., no information about how to tie the knot or what it was called was experienced during the training period). Before and after the knot training portion of the study, participants were required to complete 80 trials of a perceptual discrimination task. In this task, participants had to decide whether two photographs were of the same or different knots. Each pair of knots was selected from the same learning class (i.e., name, tie, both name and tie or untrained knots). When photographs were of the same knot, the viewing angle of the two photographs differed, so the stimuli presented were never identical. Of the 80 trials, 50\% of the knots were the same and 50\% of the knots were different. Both accuracy and response times (ms) of the perceptual discrimination task were recorded. As we were interested in modelling only the behavioural results of the perceptual discrimination task, we summarise the aspects of the dataset relevant to the present study below and refer readers to the original paper for all other details (see Cross et al., 2012).

The data included in the present paper differed from that reported by Cross and colleagues (2012) in a number of key ways. First, while 30 participants took part in the behavioural portion of the original study, only data from the 20 participants who comprised the final fMRI sample were available for re-analysis. The decision to include only the final training session in the LBA model meant that a further two participants were excluded from modelling as they did not complete the final session of the perceptual discrimination task. Therefore, 18 participants, all right-handed (Oldfield, 1971) and neurologically healthy undergraduate and graduate students, were included in the present study (M = 19.4 years, age ranged from 17 -- 27 years). We do not have access to information on the gender of participants, but we note from the original publication that of the 20 participants that completed the fMRI component of the study, 14 were female (Cross et al., 2012). This means that our sample of 18 participants is predominantly female.

Second, unlike Cross and colleagues who included the post-training session across all five days in their analysis of accuracy and RT, we included only the post-training session on the final day of training (Day 5 post-training). Given the aim of our study was to use evidence accumulation modelling to examine the cognitive mechanisms underlying linguistic and action-based training, it was sufficient to use the training effects evident on the last day of training for our analysis. That is, general training effects could be compared via a comparison between the untrained and trained conditions, and between each type of training category (e.g.~linguistic vs.~action vs.~both). The benefit of including only the final post-training sessions was a computationally simpler model.

@@

Similarly, while we had originally planned to model the difference between the pre- and post-training data for each participant, due to model complexity we were limited to including the post-training session only. It is possible, therefore, that improved post-training scores in the trained conditions, rather than reflecting the operation of training, may instead be due to unbalanced pre-test scores between these conditions. However, it should be noted that there were no clear and obvious differences in RT or accuracy in the day 1, session 1 data (i.e., before any training had started; see supplementary materials). Therefore, we consider it quite unlikely that random variation in pre-training differences between conditions could account for our findings.

In using secondary data analysis in the present study, we were also limited in what data was available for analysis. Indeed, we determined our sample size by aiming to use as much of the existing data as was possible. Consequently, the likely precision of our parameter estimates is unknown. Given this limitation, it's important for future studies look to replicate the findings in the present study.

### 2025-02-03 manuscript/supplementary/supplementary.Rmd
updated the manuscript and supps to reflect editor feedback

@@

Plots of prior and posterior distributions revealed strong updating (i.e., posteriors dominated priors), making it clear that the prior assumptions had little influence on posterior estimates (see Supplementary Figure 2).

@@

Supplementary Figures 3 displays the fits of the LBA model to the data in terms of defective cumulative distribution functions (lines) and 10th, 30th, 50th, 70th and 90th percentiles (points from left to right) averaged over participants. Thick black lines and open points correspond to the data and the thin grey lines and solid black points correspond to the model predictions averaged over posterior samples. The grey points correspond to percentile predictions for 100 randomly selected sets of posterior parameter samples, so their spread gives an idea of the uncertainty in the model’s predictions. As can be seen from the Supplementary Figure 3, the average fit of the selected LBA model was reasonable.

>>>>

### Day 1, Session 1 Data ###

We also visually checked for any clear and obvious differences between training conditions that could have existed before any training had occurred (see Supplementary Figure 2). 
In other words, we checked for random pre-training differences in RT and accuracy that might have existed in the Day 1, Session 1 data.

(ref:day1-caption) Accuracy and Reaction Time by Training Type for the Day 1 and Session 1 data (i.e., before any training had occurred). (A) Percent correct on the perceptual discrimination task by training type. (B) Mean reaction time (ms) on the perceptual discrimination task by training type. Black points denote group means, while error bars represent within subjects’ standard errors of the mean. Grey points denote individual subject means.

```{r day1-plot, fig.cap = "(ref:day1-caption)"}
knitr::include_graphics("../../figures/descriptive/day1_session1_violin_acc_rt.jpeg")
```

At least in terms of accuracy and RT when analysed separately, there appears to be no clear and obvious pre-training differences between the conditions. Moreover, for RTs, any small numerical differences between conditions are in the opposite direction to the effect of training that we observed in RT, which is that trained conditions became faster than the untrained condition. Therefore, it seems unlikely that pre-existing, chance differences between conditions could account for our findings.

@@

Plots of prior and posterior distributions revealed strong updating (i.e., posteriors dominated priors), making it clear that the prior assumptions had little influence on posterior estimates (see Supplementary Figure 3).

@@

Supplementary Figures 4 displays the fits of the LBA model to the data in terms of defective cumulative distribution functions (lines) and 10th, 30th, 50th, 70th and 90th percentiles (points from left to right) averaged over participants. Thick black lines and open points correspond to the data and the thin grey lines and solid black points correspond to the model predictions averaged over posterior samples. The grey points correspond to percentile predictions for 100 randomly selected sets of posterior parameter samples, so their spread gives an idea of the uncertainty in the model’s predictions. As can be seen from the Supplementary Figure 4, the average fit of the selected LBA model was reasonable.

### 2025-02-03 manuscript/supplementary/supplementary.tex
updated the manuscript and supps to reflect editor feedback

@@

Supplementary Figures 3 displays the fits of the LBA model to the data in terms of defective cumulative distribution functions (lines) and 10th, 30th, 50th, 70th and 90th percentiles (points from left to right) averaged over participants. Thick black lines and open points correspond to the data and the thin grey lines and solid black points correspond to the model predictions averaged over posterior samples. The grey points correspond to percentile predictions for 100 randomly selected sets of posterior parameter samples, so their spread gives an idea of the uncertainty in the model's predictions. As can be seen from the Supplementary Figure 3, the average fit of the selected LBA model was reasonable.

>>>>

\subsubsection{Day 1, Session 1 Data}\label{day-1-session-1-data}

We also visually checked for any clear and obvious differences between training conditions that could have existed before any training had occurred (see Supplementary Figure 2).
In other words, we checked for random pre-training differences in RT and accuracy that might have existed in the Day 1, Session 1 data.

{\centering \includegraphics[height=0.67\textheight,]{../../figures/descriptive/day1_session1_violin_acc_rt}

\caption{Accuracy and Reaction Time by Training Type for the Day 1 and Session 1 data (i.e., before any training had occurred). (A) Percent correct on the perceptual discrimination task by training type. (B) Mean reaction time (ms) on the perceptual discrimination task by training type. Black points denote group means, while error bars represent within subjects' standard errors of the mean. Grey points denote individual subject means.}\label{fig:day1-plot}
\end{figure}

At least in terms of accuracy and RT when analysed separately, there appears to be no clear and obvious pre-training differences between the conditions. Moreover, for RTs, any small numerical differences between conditions are in the opposite direction to the effect of training that we observed in RT, which is that trained conditions became faster than the untrained condition. Therefore, it seems unlikely that pre-existing, chance differences between conditions could account for our findings.

@@

Supplementary Figures 4 displays the fits of the LBA model to the data in terms of defective cumulative distribution functions (lines) and 10th, 30th, 50th, 70th and 90th percentiles (points from left to right) averaged over participants. Thick black lines and open points correspond to the data and the thin grey lines and solid black points correspond to the model predictions averaged over posterior samples. The grey points correspond to percentile predictions for 100 randomly selected sets of posterior parameter samples, so their spread gives an idea of the uncertainty in the model's predictions. As can be seen from the Supplementary Figure 4, the average fit of the selected LBA model was reasonable.
